Online Sequential Monte Carlo smoother for partially observed diffusion processes

### 2017-09-06 TexSubmission/ParisEM.tex
Relu jusque page 7, à finir

\section{The Generalized Random PaRIS algorithm}
\label{sec:rwparis}
%. These estimators are designed for stochastic processes
Let $(X_t)_{t\ge 0}$ be defined as a weak solution to the following SDE in $\mathbb{R}^d$:
\begin{equation}
\label{eq:target:sde}
X_0 = x_0\quad\mbox{and}\quad \rmd X_t = \alpha(X_t)\rmd t + \Gamma(X_t)\rmd W_t\eqsp,
\end{equation}
where $(W_t)_{t\ge 0}$ is a standard Brownian motion on $\mathbb{R}^d$, $\alpha: \mathbb{R}^d\to\mathbb{R}^d$ and $\Gamma: \mathbb{R}^d\to \mathbb{R}^{d\times d}$. 
In this paper, we focus on unbiased estimator of the unknown transition densities given by generalized Poisson estimators defined in \cite{fearnhead:papaspiliopoulos:roberts:2008}.
 The use of generalized Poisson estimators requires that the diffusion coefficient $\Gamma$ is constant and equals to the identity matrix.
  Note that, when this is not the case, there exists a class of diffusion processes that may be transformed to satisfy this assumption. 
  This paper then focuses on reducible SDE for which there exists an invertible and infinitely differentiable function $\eta$ such that the process $\left\{Z_t = \eta(X_t)\right\}_{t\ge 0}$ satisfies the SDE  $z_0 = \eta(x_0)$ and
\begin{equation}
\label{eq:diff:lamperti}
\mathrm{d}Z_t = \beta(Z_t)\mathrm{d}t + \mathrm{d}W_t\eqsp.
\end{equation}
By Ito's formula,  it is straightforward to show that, in the case of a reducible diffusion, the Jacobian matrix of $\eta$ satisfies
\[
\nabla \eta = \Gamma^{-1}\eqsp,
\]
and, in the case $d=1$,
\[
\beta:u \mapsto \frac{\alpha(\eta^{-1}(u))}{\Gamma(\eta^{-1}(u))} - \frac{\Gamma'(\eta^{-1}(u))}{2}\eqsp.
\]
In the case of a scalar diffusion, this Lamperti transform is given by:
\[
\eta:x\mapsto \int_{x_0}^x\Gamma^{-1}(u)\mathrm{d}u\eqsp.
\]
In the general case, \cite[Proposition~$1$]{ait-sahalia:2008} shows that when $\Gamma$ is nonsingular, the SDE is reducible if and only if, for all $1\le i,j,k\le d$,
\begin{equation}
\label{eq:reducibility}
\frac{\partial\Gamma^{-1}_{i,j}}{\partial x_k} = \frac{\partial\Gamma^{-1}_{i,k}}{\partial x_j}\eqsp.
\end{equation}
In the case of a diagonal matrix $\Gamma$ \eqref{eq:reducibility} is equivalent to assume that $\Gamma$ is such that for each $1\le i\le d$, $\Gamma_{i,i}$ depends on $x_i$ only. \cite{ait-sahalia:2008} notes that the reducibility condition \eqref{eq:reducibility} holds also for some nondiagonal matrices $\Gamma$.
 This is true in particular in the case $d=2$ for stochastic volatility models where $\sigma$ is of the form:
\[
\Gamma(x) = \begin{pmatrix}a(x_1) & a(x_1)b(x_2)\\0&c(x_2)\end{pmatrix}\eqsp.
\]
In the following, it is assumed that the process $(X_t)_{t\ge 0}$ satisfies the SDE \eqref{eq:target:sde} with $\Gamma$ being the identity matrix, i.e. we consider a diffusion after the application of the Lamperti transform.
 In addition, designing general Poisson estimators also requires that:
\begin{enumerate}[i)]
\item $\alpha$ is of the form $\alpha(x) = \nabla_x A(x)$ where $A: \mathbb{R}^d \to \mathbb{R}$ is a twice continuously differentiable function ;
\item the function $x\mapsto \left(\|\alpha(x)\|^2  + \triangle A(x)\right)/2$ is lower bounded where $\triangle$ is the Laplace operator.
\end{enumerate}
Assumption i) is somewhat restrictive as it requires $\alpha$ to derive from a scalar potential, however, it has natural applications in many fields, for instance in movement ecology, see \cite{gloaguen:etienne:lecorff:2017}. Assumption ii) is technical {\color{red} JAI LIMPRESSION QUE CEST TOUJOURS VRAI SI A(X) EST DEFINI SUR $R^d$}.

@@

The solution to \eqref{eq:target:sde} is supposed to be partially observed at times $t_0=0,\dots,t_n$ through an observation process $(Y_k)_{0\le k \le n}$ in $(\mathbb{R}^m)^{n+1}$. 
For all $0\le k \le n$, the distribution of $Y_k$ given $X_k:= X_{t_k}$ has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g(X_k,\cdot) = g_k(X_k)$. 
The distribution of $X_0$ has a density with respect to a reference measure $\mu$ on $\mathbb{R}^d$ given by $\chi$.
For all $0\le k \le n-1$, the conditional distribution of $X_{k+1} $ given $X_{k}$ has a density $\qk(X_{k},\cdot)$ with respect to $\mu$.

Let $0 \leq k \leq k' \leq n$, the joint smoothing distributions of the hidden states are defined, for all measurable function $h$ on $(\mathbb{R}^d)^{k'-k + 1}$, by:
\[
\phi_{k:k'|n}[h] = \mathbb{E}\left[h(X_k,\ldots,X_{k'})\middle|Y_{0:n}\right]\eqsp.
\]
For all $0\le k\le n$, $\phi_{k} = \phi_{k:k|k}$ denote the filtering distributions. The aim of this section is to detail the extension of PaRIS algorithm to approximate expectations of the form
\begin{equation}
\label{def:addfunc}
\phi_{0:n\vert n}[H_{n}] = \mathbb{E}\left[H_n(X_{0:n})\middle|Y_{0:n}\right] \text{ where } H_n=\sum_{k=0}^{n-1}h_k(X_k,X_{k+1})\eqsp,
\end{equation}
when the transition density of the hidden states is not available explicitly and where $\{h_k\}_{k=0}^{n-1}$ are given functions on $\mathbb{R}^d\times \mathbb{R}^d$. 
Smoothed additive functionals as \eqref{def:addfunc} are crucial for maximum likelihood inference of latent data models. 
These quantities appear naturally when computing the Fisher score in hidden Markov models or the intermediate quantity of the Expectation Maximization algorithm (see Section~\ref{sec:exp}). 
They are also pivotal to design online Expectation Maximization based algorithms which motivates the method introduced in this paper that does not require growing storage and can process observations online.

@@

At time $k = 0$, $N$ particles $\{\xi^{\ell}_0\}_{\ell=1}^N$ are sampled independently according to  $\xi^{\ell}_0 \sim \eta_0$, where $\eta_0$ is a probability density with respect to $\mu$. 
Then, $\xi^{\ell}_0$ is associated with the importance weights $\omega_0^{\ell} = \chi(\xi^{\ell}_0)g_0 (\xi^{\ell}_0)/\eta_0(\xi^{\ell}_0)$. 
For any bounded and measurable function $h$ defined on $\mathbb{R}^d$, the expectation $\phi_{0}[h] $ is approximated by
\[
\phi^N_{0}[h] = \frac{1}{\Omega_0^N} \sum_{\ell=1}^N \omega_0^{\ell} h \left(\xi^{\ell}_0 \right)\eqsp, \quad \Omega_0^N:= \sum_{\ell=1}^N \omega_0^{\ell}\eqsp.
\]
Then, for $1\le k \le n$, using $\{(\xi^{\ell}_{k-1},\omega^{\ell}_{k-1})\}_{\ell=1}^N$, the auxiliary particle filter of \cite{pitt:shephard:1999} samples pairs $\{(I^{\ell}_k,\xi^{\ell}_{k})\}_{\ell=1}^N$ of indices and particles using an instrumental transition density $p_{k-1}$ on $\mathbb{R}^d\times \mathbb{R}^d$ and an adjustment multiplier function $\vartheta_k$ on $\mathbb{R}^d$.
 Each new particle $\xi^{\ell}_{k}$ and weight $\omega^{\ell}_k$ at time $k$ are computed following these steps:
\begin{enumerate}[-]
\item choose a particle index $I^{\ell}_k$ at time $k-1$ in $\{1,\ldots,N\}$ with probabilities proportional to $\omega_{k-1}^{j} \vartheta_k (\xi^{j}_{k-1})$, for $j$ in $\{1,\ldots,N\}$ ;
\item sample  $\xi^{\ell}_{k}$ using this chosen particle according to $\xi^{\ell}_{k} \sim p_{k-1}(\xi^{I^{\ell}_k}_{k-1},\cdot)$ ; 
\item  associate the particle $\xi^{\ell}_k$ with the importance weight:
\begin{equation}
\label{eq:importance:weights}
\omega^{\ell}_k := \frac{q_{k-1}(\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)g_k(\xi^{\ell}_k)}{\vartheta_k(\xi^{I^{\ell}_k}_{k-1}) p_{k-1} (\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)}\eqsp.
\end{equation}
\end{enumerate} 
The expectation $\phi_{k}[h]$ is approximated by
\[
\phi^N_{k}[h] := \frac{1}{\Omega_k^N} \sum_{\ell=1}^N \omega_k^{\ell} h \left(\xi^{\ell}_k \right)\eqsp,\quad\Omega_k^N:= \sum_{\ell=1}^N \omega_k^{\ell}\eqsp.
\]
The most simple choice for $p_{k-1}$ and $\vartheta_k$ is the bootstrap filter proposed by \cite{gordon:salmond:smith:1993} which sets $p_{k-1} = q_{k-1}$ and for all $x\in\mathbb{R}^d$, $\vartheta(x) = 1$.
 In the case of POD processes $q_{k-1}$ is unknown but it can be replaced by any approximation to sample the particles as any choice of $p_{k-1}$ can be made.
  The approximation can be obtained using a discretization scheme such as Euler method or a Poisson based approximation as detailed below.
   A more appealing choice is the fully adapted particle filter which sets for all $x,x'\in\mathbb{R}^d$, $p_{k-1}(x,x') \propto q_{k-1}(x,x')g_k(x')$ and for all $x\in\mathbb{R}^d$, $\vartheta(x) = \int q_{k-1}(x,x')g_k(x')\mu(\rmd x')$.
   Here again $q_{k-1}$ has to be replaced by an approximation. In Section~\ref{sec:exp}, it is replaced by the Gaussian approximation provided by a Euler scheme which leads to a Gaussian proposal density $p_{k-1}$ as the observation model is linear and Gaussian.

>>>>

\section{The Generalized Random PaRIS algorithm}
\label{sec:rwparis}
%. These estimators are designed for stochastic processes
Let $(X_t)_{t\ge 0}$ be defined as a weak solution to the following SDE in $\mathbb{R}^d$:
\begin{equation}
\label{eq:target:sde}
X_0 = x_0\quad\mbox{and}\quad \rmd X_t = \alpha(X_t)\rmd t + \Gamma(X_t)\rmd W_t\eqsp,
\end{equation}
where $(W_t)_{t\ge 0}$ is a standard Brownian motion on $\mathbb{R}^d$, $\alpha: \mathbb{R}^d\to\mathbb{R}^d$ and $\Gamma: \mathbb{R}^d\to \mathbb{R}^{d\times d}$. 
The algorithm proposed in this paper requires an  unbiased estimator of the unknown transition densities and we propose to use  generalized Poisson estimators defined in \cite{fearnhead:papaspiliopoulos:roberts:2008}.
 The use of generalized Poisson estimators entails that the diffusion coefficient $\Gamma$ is constant and equals to the identity matrix.
  Note that, when this is not the case, there exists a class of diffusion processes that may be transformed using Lamperti transformation to satisfy this assumption. 
  This paper then focuses on reducible SDE for which there exists an invertible and infinitely differentiable function $\eta$ such that the process $\left\{Z_t = \eta(X_t)\right\}_{t\ge 0}$ satisfies the SDE  $z_0 = \eta(x_0)$ and
\begin{equation}
\label{eq:diff:lamperti}
\mathrm{d}Z_t = \beta(Z_t)\mathrm{d}t + \mathrm{d}W_t\eqsp.
\end{equation}
By Ito's formula,  it is straightforward to show that, in the case of a reducible diffusion, the Jacobian matrix of $\eta$ satisfies
\[
\nabla \eta = \Gamma^{-1}\eqsp,
\]
and, in the case $d=1$,
\[
\beta:u \mapsto \frac{\alpha(\eta^{-1}(u))}{\Gamma(\eta^{-1}(u))} - \frac{\Gamma'(\eta^{-1}(u))}{2}\eqsp.	
\]
In the case of a scalar diffusion, this Lamperti transform is given by:
\[
\eta:x\mapsto \int_{x_0}^x\Gamma^{-1}(u)\mathrm{d}u\eqsp.
\]
In the general case, \cite[Proposition~$1$]{ait-sahalia:2008} shows that when $\Gamma$ is nonsingular, the SDE is reducible if and only if, for all $1\le i,j,k\le d$,
\begin{equation}
\label{eq:reducibility}
\frac{\partial\Gamma^{-1}_{i,j}}{\partial x_k} = \frac{\partial\Gamma^{-1}_{i,k}}{\partial x_j}\eqsp.
\end{equation}
In the case of a diagonal matrix $\Gamma$ \eqref{eq:reducibility} is equivalent to assume that $\Gamma$ is such that for each $1\le i\le d$, $\Gamma_{i,i}$ depends on $x_i$ only. \cite{ait-sahalia:2008} notes that the reducibility condition \eqref{eq:reducibility} holds also for some nondiagonal matrices $\Gamma$.
 This is true in particular in the case $d=2$ for stochastic volatility models where $\sigma$ is of the form:
\[
\Gamma(x) = \begin{pmatrix}a(x_1) & a(x_1)b(x_2)\\0&c(x_2)\end{pmatrix}\eqsp.
\]
In the following, it is assumed that the process $(X_t)_{t\ge 0}$ satisfies the SDE \eqref{eq:target:sde} with $\Gamma$ being the identity matrix, i.e. we consider a diffusion after the application of the Lamperti transform.
 In addition, designing general Poisson estimators also requires that:
\begin{enumerate}[i)]
\item $\alpha$ is of the form $\alpha(x) = \nabla_x A(x)$ where $A: \mathbb{R}^d \to \mathbb{R}$ is a twice continuously differentiable function ;
\item the function $x\mapsto \left(\|\alpha(x)\|^2  + \triangle A(x)\right)/2$ is lower bounded where $\triangle$ is the Laplace operator.
\end{enumerate}
Assumption i) is somewhat restrictive as it requires $\alpha$ to derive from a scalar potential, however, it has natural applications in many fields, for instance in movement ecology, see \cite{gloaguen:etienne:lecorff:2017}. Assumption ii) is technical {\color{red} JAI LIMPRESSION QUE CEST TOUJOURS VRAI SI A(X) EST DEFINI SUR $R^d$}.{\color{blue} pkoi  $\triangle A(x)$ serait il borné inférieurement ?}

@@

The solution to \eqref{eq:target:sde} is supposed to be partially observed at times $t_0=0,\dots,t_n$ through an observation process $(Y_k)_{0\le k \le n}$ in $(\mathbb{R}^m)^{n+1}$. 
For all $0\le k \le n$, the distribution of $Y_k$ given $X_k$, $X_k$ being a short noation for $X_{t_k}$, has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g(X_k,\cdot)$. For the sake of silmplicity, we will use the classical short notation $g_k(X_k)$ instead of  $g(X_k,\cdot)$. 
The distribution of $X_0$ has a density with respect to a reference measure $\mu$ on $\mathbb{R}^d$ given by $\chi$.
For all $0\le k \le n-1$, the conditional distribution of $X_{k+1} $ given $X_{k}$ has a density $\qk(X_{k},\cdot)$ with respect to $\mu$.

Let $0 \leq k \leq k' \leq n$, the joint smoothing distributions of the hidden states are defined, for all measurable function $h$ on $(\mathbb{R}^d)^{k'-k + 1}$, by:
\[
\phi_{k:k'|n}[h] = \mathbb{E}\left[h(X_k,\ldots,X_{k'})\middle|Y_{0:n}\right]\eqsp.
\]
For all $0\le k\le n$, $\phi_{k} = \phi_{k:k|k}$ denotes the filtering distributions. The aim of this section is to detail the extension of PaRIS algorithm to approximate expectations of the form
\begin{equation}
\label{def:addfunc}
\phi_{0:n\vert n}[H_{n}] = \mathbb{E}\left[H_n(X_{0:n})\middle|Y_{0:n}\right] \text{ where } H_n=\sum_{k=0}^{n-1}h_k(X_k,X_{k+1})\eqsp,
\end{equation}
when the transition density of the hidden states is not available explicitly and where $\{h_k\}_{k=0}^{n-1}$ are given functions on $\mathbb{R}^d\times \mathbb{R}^d$. 
Smoothed additive functionals as \eqref{def:addfunc} are crucial for maximum likelihood inference of latent data models. 
These quantities appear naturally when computing the Fisher score in hidden Markov models or the intermediate quantity of the Expectation Maximization algorithm (see Section~\ref{sec:exp}). 
They are also pivotal to design online Expectation Maximization based algorithms which motivates the method introduced in this paper that does not require growing storage and can process observations online.

@@

At time $k = 0$, $N$ particles $\{\xi^{\ell}_0\}_{\ell=1}^N$ are sampled independently according to  $\xi^{\ell}_0 \sim \eta_0$, where $\eta_0$ is a probability density with respect to $\mu$. 
Then, $\xi^{\ell}_0$ is associated with the importance weights $\omega_0^{\ell} = \chi(\xi^{\ell}_0)g_0 (\xi^{\ell}_0)/\eta_0(\xi^{\ell}_0)$. 
For any bounded and measurable function $h$ defined on $\mathbb{R}^d$, the expectation $\phi_{0}[h] $ is approximated by
\[
\phi^N_{0}[h] = \frac{1}{\Omega_0^N} \sum_{\ell=1}^N \omega_0^{\ell} h \left(\xi^{\ell}_0 \right)\eqsp, \quad \Omega_0^N:= \sum_{\ell=1}^N \omega_0^{\ell}\eqsp.
\]
Then, for $1\le k \le n$, using $\{(\xi^{\ell}_{k-1},\omega^{\ell}_{k-1})\}_{\ell=1}^N$, the auxiliary particle filter of \cite{pitt:shephard:1999} samples pairs $\{(I^{\ell}_k,\xi^{\ell}_{k})\}_{\ell=1}^N$ of indices and particles using an instrumental transition density $p_{k-1}$ on $\mathbb{R}^d\times \mathbb{R}^d$ and an adjustment multiplier function $\vartheta_k$ on $\mathbb{R}^d$.
 Each new particle $\xi^{\ell}_{k}$ and weight $\omega^{\ell}_k$ at time $k$ are computed following these steps:
\begin{enumerate}[-]
\item choose a particle index $I^{\ell}_k$ at time $k-1$ in $\{1,\ldots,N\}$ with probabilities proportional to $\omega_{k-1}^{j} \vartheta_k (\xi^{j}_{k-1})$, for $j$ in $\{1,\ldots,N\}$ ;
\item sample  $\xi^{\ell}_{k}$ using this chosen particle according to $\xi^{\ell}_{k} \sim p_{k-1}(\xi^{I^{\ell}_k}_{k-1},\cdot)$ ; 
\item  associate the particle $\xi^{\ell}_k$ with the importance weight:
\begin{equation}
\label{eq:importance:weights}
\omega^{\ell}_k := \frac{q_{k-1}(\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)g_k(\xi^{\ell}_k)}{\vartheta_k(\xi^{I^{\ell}_k}_{k-1}) p_{k-1} (\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)}\eqsp.
\end{equation}
\end{enumerate} 
The expectation $\phi_{k}[h]$ is approximated by
\[
\phi^N_{k}[h] := \frac{1}{\Omega_k^N} \sum_{\ell=1}^N \omega_k^{\ell} h \left(\xi^{\ell}_k \right)\eqsp,\quad\Omega_k^N:= \sum_{\ell=1}^N \omega_k^{\ell}\eqsp.
\]
The most simple choice for $p_{k-1}$ and $\vartheta_k$ is the bootstrap filter proposed by \cite{gordon:salmond:smith:1993} which sets $p_{k-1} = q_{k-1}$ and for all $x\in\mathbb{R}^d$, $\vartheta(x) = 1$.
 In the case of POD processes $q_{k-1}$ is unknown but it can be replaced by any approximation to sample the particles as any choice of $p_{k-1}$ can be made.
  The approximation can be obtained using a discretization scheme such as Euler method or a Poisson based approximation as detailed below.
   A more appealing choice is the fully adapted particle filter which sets for all $x,x'\in\mathbb{R}^d$, $p_{k-1}(x,x') \propto q_{k-1}(x,x')g_k(x')$ and for all $x\in\mathbb{R}^d$, $\vartheta(x) = \int q_{k-1}(x,x')g_k(x')\mu(\rmd x')$. {\color{blue} C'est si clair que $\vartheta$ est calculable pratiquement ?}
   Here again $q_{k-1}$ has to be replaced by an approximation. In Section~\ref{sec:exp}, it is replaced by the Gaussian approximation provided by a Euler scheme which leads to a Gaussian proposal density $p_{k-1}$ as the observation model is linear and Gaussian.
